---
title: "Getting started with hypnoR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with hypnoR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 8,
  fig.height = 3.5,
  dpi        = 150,
  out.width  = "100%"
)
has_mrpheus <- requireNamespace("mrpheus", quietly = TRUE)
```

```{r packages, message = FALSE}
library(hypnoR)
```

## Overview

**hypnoR** is the hypnogram layer of the Circadia Lab R ecosystem. It takes a
staged hypnogram from anywhere -- **mrpheus** (full AASM: `W` / `N1` / `N2` /
`N3` / `REM`), **zeitR** (coarse actigraphy-derived: `W` / `Sleep` / `Quiet
sleep`), or a hand-built tibble -- and turns it into sleep architecture
metrics, NREM/REM cycles, transition statistics, and publication-ready plots.

Every metric and plotting function is **staging-agnostic**: it computes
whatever is possible given the resolution actually present, and returns `NA`
(documented per-function) for metrics that need finer staging than what's
available.

This article walks through the whole pipeline using a real recording:
mrpheus's bundled `SC4001E0` example -- a 22-hour Sleep-EDF cassette
recording, already staged by mrpheus's automatic sleep-staging model.

```{r check-mrpheus, echo = FALSE, results = "asis"}
if (!has_mrpheus) {
  cat(
    "_mrpheus isn't installed in the environment that built this article, ",
    "so the code below is shown but wasn't executed. Install mrpheus to ",
    "see it run._",
    sep = ""
  )
}
```

## Installation

```r
# install.packages("remotes")
remotes::install_github("circadia-bio/hypnoR")
```

## From mrpheus to hypnoR

```{r load-staging, eval = has_mrpheus}
staging <- readRDS(system.file("extdata", "SC4001E0_staging.rds", package = "mrpheus"))
staging
```

`staging` is mrpheus's raw per-epoch output: one row per 30-second epoch, an
AASM `stage` label, and five posterior probabilities from the underlying
LightGBM classifier. `mrpheus::export_hypnogram()` prepares it for hypnoR:

```{r export-hypnogram, eval = has_mrpheus}
mrp_hyp <- mrpheus::export_hypnogram(
  staging,
  epoch_s        = 30,
  start_time     = as.POSIXct("2024-01-01 16:13:00", tz = "UTC"),
  participant_id = "SC4001"
)
```

(The real recording start time comes from the EDF header at
`mrpheus::read_edf()` time; the bundled `.rds` doesn't carry it, so we use a
placeholder here based on the note in mrpheus's own `sleep-staging-demo`
article that the cassette started around 16:13 local time.)

`new_hypnogram()` -- the constructor every other hypnoR function goes
through -- accepts this directly:

```{r new-hypnogram, eval = has_mrpheus}
hyp <- new_hypnogram(mrp_hyp)
hyp
```

Staging resolution (`aasm` here, since `N1` / `N2` / `N3` / `REM` are
present) and epoch duration are auto-detected and stored as attributes on
`hyp` -- every other hypnoR function reads them from there rather than
asking you to repeat yourself.

## Preprocessing

Two optional cleanup steps happen before any metric sees the hypnogram.

### Smoothing

Automatic per-epoch staging with no temporal continuity constraint -- as
here -- can produce brief, isolated stage flips that don't reflect real sleep
physiology (a single REM epoch nested inside an N2 run, for instance).
`smooth_hypnogram()` offers two label-only rules:

```{r smooth, eval = has_mrpheus}
hyp_smooth <- smooth_hypnogram(hyp, method = c("aasm_isolated", "min_run"), min_run_epochs = 4)
mean(hyp_smooth$stage != hyp_smooth$stage_raw)
```

`aasm_isolated` reassigns a single epoch flanked identically on both sides;
`min_run` merges any run shorter than `min_run_epochs` into whichever
flanking run is longer, regardless of whether the flanks agree. The original,
unsmoothed labels are always preserved in `stage_raw` for comparison.

### Windowing

`SC4001` is a 22-hour ambulatory recording -- most of it is ordinary daytime
Wake before and after the one real sleep period. Running metrics over the
*entire* recording would badly distort things like sleep onset latency and
NREM/REM cycle counts, so `window_hypnogram()` restricts a hypnogram to a
time or epoch window before any metric sees it:

```{r window, eval = has_mrpheus}
sleep_idx    <- which(as.character(hyp_smooth$stage) != "W")
onset_epoch  <- hyp_smooth$epoch[sleep_idx[1]]
offset_epoch <- hyp_smooth$epoch[sleep_idx[length(sleep_idx)]]

hyp_sleep <- window_hypnogram(hyp_smooth, from_epoch = onset_epoch, to_epoch = offset_epoch)
nrow(hyp_sleep)
```

`compute_sleep_architecture()` also accepts `lights_off`/`lights_on`
directly and windows internally, if you'd rather pass real lights-off/on
timestamps than derive a window from the data itself.

## Sleep architecture

```{r architecture, eval = has_mrpheus}
arch <- compute_sleep_architecture(hyp_sleep)
arch
```

## Stage transitions

```{r transitions, eval = has_mrpheus}
trans <- compute_transitions(hyp_sleep)
trans$matrix
trans$fragmentation
```

## NREM/REM cycles

```{r cycles, eval = has_mrpheus}
cyc <- compute_cycles(hyp_sleep, method = "aasm")
cyc
```

## Plotting

Two hypnogram styles are available. `"step"` (the default) is the classic
clinical trace:

```{r plot-hypnogram-step, eval = has_mrpheus, fig.height = 3}
plot_hypnogram(hyp_sleep, cycles = cyc)
```

`"capsule"` draws rounded-pill bars per contiguous stage run, one lane per
stage:

```{r plot-hypnogram-capsule, eval = has_mrpheus, fig.height = 3}
plot_hypnogram(hyp_sleep, style = "capsule")
```

```{r plot-architecture, eval = has_mrpheus, fig.height = 3.5}
plot_architecture(arch)
```

```{r plot-transitions, eval = has_mrpheus, fig.height = 4}
plot_transition_matrix(trans$matrix)
```

## What isn't here yet

`read_hypnogram()` -- ingesting hypnograms directly from CSV, EDF
annotations, YASA, Compumedics, or Nox export files -- is still under
development. For now, hypnograms come from an upstream package
(`mrpheus::export_hypnogram()`, `zeitR::export_hypnogram()`) or a hand-built
tibble passed to `new_hypnogram()`, as shown above.

See `vignette("mrpheus-integration", package = "hypnoR")` (or the "Worked
examples" section of the pkgdown site) for a deeper, warts-and-all walk
through cleaning up and interpreting real automatic staging output.
