---
title: "Coarse hypnograms from actigraphy: a worked example with zeitR"
description: |
  A walkthrough of taking zeitR's rest-activity pipeline output through
  hypnoR's coarse-staging path, using zeitR's own bundled ActTrust
  validation recording. Along the way: timestamp gaps, why off-wrist time
  quietly becomes "Wake," picking the right night out of a multi-day
  recording, and whether smoothing even matters for actigraphy-derived
  staging the way it did for mrpheus's raw automatic staging.
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse   = TRUE,
  comment    = "#>",
  fig.width  = 9,
  fig.height = 3.5,
  dpi        = 150,
  out.width  = "100%"
)
has_zeitr <- requireNamespace("zeitR", quietly = TRUE)
```

```{r packages, message = FALSE}
library(hypnoR)
```

```{r check-zeitr, echo = FALSE, results = "asis"}
if (!has_zeitr) {
  cat(
    "_zeitR isn't installed in the environment that built this article, ",
    "so the code below is shown but wasn't executed._",
    sep = ""
  )
}
```

The [`mrpheus-integration`](mrpheus-integration.html) article walks through
hypnoR's AASM path end-to-end using mrpheus's raw automatic PSG staging.
Actigraphy-derived staging comes with a different set of things worth
checking before you trust the metrics -- this picks up that same "check
before you interpret" habit, using zeitR's bundled ActTrust recording
instead.

## Getting the data in

zeitR ships a real validation recording, `input1.txt`, used in its own
pipeline-parity tests against a Python reference implementation:

```{r run-pipeline, eval = has_zeitr, message = FALSE}
result <- zeitR::run_pipeline(
  system.file("extdata", "input1.txt", package = "zeitR"),
  tz    = "UTC",
  quiet = TRUE
)
result
```

The `quiet = TRUE` above is deliberate, not just noise-suppression -- worth
checking what it's actually quieting rather than reflexively silencing it:

```{r check-consistency, eval = has_zeitr}
result$issues
```

A handful of minor timestamp gaps, no backward jumps, no year artefacts.
Small gaps like this are common in real wearable recordings (a missed
transmission, a brief disconnect) and don't necessarily invalidate the
recording -- but it's worth actually looking at `$issues` rather than
assuming `quiet = TRUE` means "nothing to see here."

## Into hypnoR

```{r export, eval = has_zeitr}
zeitr_hyp  <- zeitR::export_hypnogram(result)
hyp        <- new_hypnogram(zeitr_hyp)
attr(hyp, "resolution")
nrow(hyp)
```

Coarse resolution, auto-detected correctly (no `N1`/`N2`/`N3`/`REM` labels
in the data). `nrow(hyp)` is likely well beyond one night's worth of
epochs -- `input1.txt` is a multi-day recording, which matters for
everything that follows.

## A caveat: off-wrist time is folded into "Wake"

`zeitR::export_hypnogram()`'s stage mapping sends `state == 4` (off-wrist)
to `"W"`, the same label as genuine wakefulness -- there's no fourth
"unknown/off-wrist" category in hypnoR's 3-state coarse model. That's a
reasonable simplification, not a bug, but it means a recording with
substantial off-wrist time will show inflated Wake duration that isn't
really about sleep/wake behaviour at all. Worth checking how much of this
recording that actually is:

```{r offwrist-check, eval = has_zeitr}
offwrist_epochs <- sum(result$data$state == 4L)
offwrist_epochs
offwrist_epochs / nrow(result$data)  # proportion of the whole recording
```

If that proportion is small, it's a non-issue here. If a given analysis
turned up unexpectedly long Wake/WASO on a specific night, this is the
first place to check before concluding anything about sleep quality --
the device could simply have been off the wrist.

## Picking the right night

`input1.txt` spans several nights (`result$nights`), so running metrics
over the *entire* multi-day recording would be exactly the kind of mistake
covered in the mrpheus article: `TST` would represent multiple nights'
sleep combined, and `SOL` would be measured from the very start of the
whole file rather than from any single night's actual bedtime.

```{r nights, eval = has_zeitr}
result$nights
```

Restrict to one real night using its own `bed_time`/`get_up_time` --
exactly the `lights_off`/`lights_on` window `compute_sleep_architecture()`
and `window_hypnogram()` expect:

```{r window-night, eval = has_zeitr}
night1 <- result$nights[!result$nights$is_nap, ][1L, ]
night1

hyp_night1 <- window_hypnogram(hyp, lights_off = night1$bed_time, lights_on = night1$get_up_time)
nrow(hyp_night1)
```

## Does smoothing matter here the way it did for mrpheus?

The mrpheus article found smoothing genuinely necessary, because
`mrpheus::stage_epochs()` is a raw per-epoch argmax classifier with no
continuity constraint. zeitR's pipeline is different: Crespo (2012)'s
sleep-period detection and Cole-Kripke (1992)'s epoch scoring both impose
their own temporal structure on the output, rather than staging each epoch
independently. Worth checking whether `smooth_hypnogram()` actually finds
anything to do here, rather than assuming the same lesson transfers
automatically:

```{r smooth-check, eval = has_zeitr}
hyp_smooth <- smooth_hypnogram(hyp_night1, method = c("aasm_isolated", "min_run"), min_run_epochs = 4)
mean(hyp_smooth$stage != hyp_smooth$stage_raw)
```

Whatever that number comes out to, compare it against the mrpheus
article's equivalent check -- if it's much lower here, that's evidence the
two staging sources genuinely differ in how much raw epoch-level noise
they produce, not just a coincidence of this particular recording.

## Metrics on the properly-windowed night

```{r architecture, eval = has_zeitr}
arch_night1 <- compute_sleep_architecture(hyp_night1)
arch_night1
```

```{r transitions, eval = has_zeitr}
trans_night1 <- compute_transitions(hyp_night1)
trans_night1$matrix
trans_night1$fragmentation
```

`compute_cycles()` remains AASM-only regardless of staging source -- coarse
hypnograms have no REM stage to segment NREM/REM cycles on:

```{r cycles-error, eval = has_zeitr, error = TRUE}
compute_cycles(hyp_night1)
```

## Both plotting styles

```{r plot-step, eval = has_zeitr, fig.height = 3}
plot_hypnogram(hyp_night1)
```

```{r plot-capsule, eval = has_zeitr, fig.height = 3}
plot_hypnogram(hyp_night1, style = "capsule")
```

```{r plot-architecture, eval = has_zeitr, fig.height = 3.5}
plot_architecture(arch_night1)
```

## Main sleep vs naps, if there are any

`result$nights$is_nap` distinguishes zeitR's main sleep periods from
secondary (nap) periods. Worth checking whether this recording has any naps
at all before assuming a single "the sleep period" exists per day:

```{r naps-check, eval = has_zeitr}
table(result$nights$is_nap)
```

If there are naps present, comparing their architecture against a main
night (rather than pooling them together) avoids quietly averaging two
physiologically different kinds of sleep opportunity into one number.

## Takeaways

- Actigraphy-derived hypnograms carry different caveats than automatic PSG
  staging does -- off-wrist time masquerading as Wake is the zeitR-specific
  one to check, the way scattered low-confidence REM was the mrpheus-specific
  one.
- `quiet = TRUE` (or any warning-suppression argument) is worth checking
  what it's actually suppressing, not just trusting it silently.
- Windowing to the real analysis period matters regardless of staging
  source -- a multi-night actigraphy recording needs the same care a
  multi-hour ambulatory PSG recording does.
- Smoothing's necessity is a property of the *staging source*, not a
  universal step to always apply -- checking whether it changes anything
  is more honest than assuming a lesson from one pipeline transfers
  unchanged to another.
- `compute_cycles()`'s AASM-only requirement is a genuine constraint of
  what coarse staging can support, not an arbitrary limitation -- there's
  no REM stage in a 3-state model to segment cycles on.
